---
title: "Analysis of Cyclistic Riders"
output: html_notebook
---

Installing the tidyverse and lubridate packages for data manipulation.

```{r}
library(tidyverse)
library(lubridate)
```

The funcion clean_bikes will be created to clean the data.

```{r}
clean_bikes <- function(x) {
  file <- read_csv(x)
  file_df <- as.data.frame(file)
  
  # removes duplicates
  file_df <- file_df %>% distinct() 
  
  # add a column for trip duration
  file_df$duration <- difftime(file_df$ended_at, file_df$started_at, units="mins")
  
  # filter table where there the data is incomplete - no start station, end station name
  # rideable type, and duration where it is longers than a minute and less than 12 hours
  
  file_df = file_df %>% 
    filter(!is.na(start_station_name)) %>% 
    filter(!is.na(end_station_name)) %>% 
    filter(rideable_type == 'docked_bike' | rideable_type == 'electric_bike' | rideable_type == 'classic_bike') %>%
    filter(duration > 1 & duration <= 720)
  
  # adding a column for the day of week and hour the trips started
  file_df$weekday <- weekdays(file_df$started_at)
  file_df$hour <- hour(file_df$started_at)
  
  return(file_df)
}
```

Using the clean_bikes function, we will import the trip data for all months in 2022:

```{r}

jan_df <- clean_bikes('/202201-divvy-tripdata.csv')
feb_df <- clean_bikes('/202202-divvy-tripdata.csv')
mar_df <- clean_bikes('/202203-divvy-tripdata.csv')
apr_df <- clean_bikes('/202204-divvy-tripdata.csv')
may_df <- clean_bikes('/202205-divvy-tripdata.csv')
jun_df <- clean_bikes('/202206-divvy-tripdata.csv')
jul_df <- clean_bikes('/202207-divvy-tripdata.csv')
aug_df <- clean_bikes('/202208-divvy-tripdata.csv')
sep_df <- clean_bikes('/202209-divvy-tripdata.csv')
oct_df <- clean_bikes('/202210-divvy-tripdata.csv')
nov_df <- clean_bikes('/202211-divvy-tripdata.csv')
dec_df <- clean_bikes('/202212-divvy-tripdata.csv')

```

We can combine the data by quarter:

```{r}
# combining data for each month into quarters
q1_trips_df <- rbind(jan_df, feb_df, mar_df)
q2_trips_df <- rbind(apr_df, may_df, jun_df)
q3_trips_df <- rbind(jul_df, aug_df, sep_df)
q4_trips_df <- rbind(oct_df, nov_df, dec_df)
```

Then, combine all the quarters to have the data for the year.

```{r}
all_trips_df <- rbind(q1_trips_df, q2_trips_df, q3_trips_df, q4_trips_df)
```

We will take a sample of 500 casual members and 500 annual members. Working with 4,000,000 records would be absurd.

```{r}
  set.seed(1)
  casual_sample <- all_trips_df %>% filter(member_casual == "casual") %>% sample_n(500)
  
  set.seed(1)
  member_sample <- all_trips_df %>% filter(member_casual == "member") %>% sample_n(500)

```

Table of rideable percentages (member):

```{r}
    
    memeber_rideable_count <- member_sample %>% count(rideable_type)
    data.frame(rideable_type = member_rideable_count$rideable_type, percentage = prop.table(all_trips_casual_rideable_count$n))
```

Table of rideable percentages (casual):

```{r}
    casual_rideable_count <- casual_samp %>% count(rideable_type)
    data.frame(rideable_type = casual_rideable_count$rideable_type, percentage = prop.table(all_trips_casual_rideable_count$n))
```

Table of rideable percentages (casual, excluding docked bikes):

```{r}
    all_trips_casual_rideable_count <- casual_samp %>% filter(rideable_type != 'docked_bike')  %>% count(rideable_type)
    data.frame(rideable_type = all_trips_casual_rideable_count$rideable_type, percentage = prop.table(all_trips_casual_rideable_count$n))
```

What day do annual memmbers ride during the week?

```{r}
    member_sample %>% 
      ggplot(mapping=aes(weekday)) + 
      geom_bar() + 
      ggtitle('Weekday Usage for Annual Members') + xlab('Weekday') + 
      ylab('Number of Riders') + scale_x_discrete(limits=c("Sunday","Monday","Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
```

How about casual riders?

```{r}
    casual_sample %>% 
      ggplot(mapping=aes(weekday)) + 
      geom_bar() + 
      ggtitle('Weekday Usage for Casual Members') + xlab('Weekday') + 
      ylab('Number of Riders') + scale_x_discrete(limits=c("Sunday","Monday","Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))

```

The time of day annual members ride: 

```{r}
  member_sample %>% 
    filter(weekday != "Sunday" & weekday != "Saturday") %>% 
    ggplot(mapping=aes(hour)) + 
    geom_line(aes(fill=after_stat(count)),stat="bin",binwidth=1) + 
    geom_point(aes(fill=after_stat(count)), stat="bin", binwidth=1, show.legend = FALSE) + xlab('Hour') + 
    ylab('Number of Riders') + ggtitle('Time Started During Weekday', subtitle="For annual members") + 
    geom_rect(aes(xmin=15, xmax = 19, ymin=25, ymax=50), fill ="yellow", alpha = 0.003) + 
    geom_rect(aes(xmin=6.5, xmax = 8.5, ymin=18, ymax=35), fill ="yellow", alpha = 0.003)
```

For casual riders:

```{r}
  casual_sample %>%
    filter(weekday != "Sunday" & weekday != "Saturday") %>% 
    ggplot(mapping=aes(hour)) + 
    geom_line(aes(fill=after_stat(count)),stat="bin",binwidth=1) + 
    geom_point(aes(fill=after_stat(count)), stat="bin", binwidth=1, show.legend = FALSE) + 
    xlab('Hour') + ylab('Number of Riders') + ggtitle('Time Started During Weekday', subtitle="For casual members") +
    geom_rect(aes(xmin=14, xmax = 20, ymin=15, ymax=40), fill ="yellow", alpha = 0.003)

```



